Taking Linear Logic Apart

### 2018-04-28 doc/tlla2018/main.tex
new convention: one line per paragraph

\documentclass[draft,submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{example}{Example}
\input{preamble-typing}
\usepackage[sectionbib,square,numbers,sort&compress]{natbib}
\renewcommand*{\eg}{e.g.\@\xspace}
\renewcommand*{\ie}{i.e.\@\xspace}
\usepackage{more-space}
\usepackage{setspace}
\everypar{\looseness=-1}
\renewcommand{\baselinestretch}{0.975}

@@

This syntactic mismatch has an effect on the semantics as well.
For instance, the $\beta$-reduction for output and input in \cp is
\(
\reducesto
{\cpCut{x}{\cpSend{x}{y}{P}{Q}}{\cpRecv{x}{y}{R}}}
{\cpCut{y}{P}{\cpCut{x}{Q}{R}}}
\).
Here, the parallel composition $\tm{(\piPar{P}{Q})}$ is of no relevance to this
communication, yet the rule needs to inspect it to be able to nest the name
restrictions appropriately in the resulting term.

@@

The term language of \cp is a variant of the \textpi-calculus.
The variables $\tm{x}$, $\tm{y}$, and $\tm{z}$ range over channel names.
The construct $\tm{\cpLink{x}{y}}$ links two
channels~\cite{sangiorgi1996,boreale1998}, forwarding messages received on
$\tm{x}$ to $\tm{y}$ and vice versa.
The construct $\tm{\cpCut{x}{P}{Q}}$ creates a new channel $\tm{x}$, and composes
two processes, $\tm{P}$ and $\tm{Q}$, which communicate on $\tm{x}$, in parallel.
Therefore, in $\tm{\cpCut{x}{P}{Q}}$ the name $\tm{x}$ is bound in both $\tm{P}$
and $\tm{Q}$.
In $\tm{\cpRecv{x}{y}{P}}$ and $\tm{\cpSend{x}{y}{P}{Q}}$, round brackets denote
input, square brackets denote output.
We use bound output~\cite{sangiorgi1996}, meaning that both input and output
bind a new name.
In $\tm{\cpRecv{x}{y}{P}}$ the new name $\tm{y}$ is bound in $\tm{P}$.
In $\tm{\cpSend{x}{y}{P}{Q}}$, the new name $\tm{y}$ is only bound in $\tm{P}$,
while $\tm{x}$ is only bound in $\tm{Q}$.

@@

Terms in \cp are identified up to structural congruence, which states that
links are symmetric, and parallel compositions $\tm{\cpCut{x}{P}{Q}}$ are
associative and commutative.

@@

Reductions relate processes with their reduced forms \eg, a reduction
$\reducesto{P}{Q}$ denotes that the process $\tm{P}$ can reduce to the process
$\tm{Q}$ in a single step.

@@

  Reductions are described by the smallest relation $\Longrightarrow$ on process
  terms closed under rules in \cref{fig:cp-reduction}.
  The relation $\Longrightarrow^\star$ is the reflexive, transitive closure of 

@@

In this section, we introduce Hypersequent Classical Processes (\hcp), a variant
of \cp which registers parallelism in the typing jugdgements using
hypersequents, allowing us to take apart the monolithic term constructors of \cp
(\eg $\tm{\cpSend{x}{y}{P}{Q}}$) into the corresponding \textpi-calculus term
constructs.

@@

The term language of \hcp is a variant of \cp where the term constructs have
been taken apart into \textpi-calculus primitives.

@@

One side-effect of the changes on the term level is that our structural
congruence is much more standard. It has associativity, commutativity, and a
unit for parallel composition, commutativity of name restrictions, and scope
extrusion.

@@

@@

There are two changes to the reduction system.
First, because $\tm{\piBoundSend{x}{y}{P}}$ and $\tm{\piBoundSend{x}{}{P}}$ are
now terms in their own right, the $\cpRedBetaTensParr$ and $\cpRedBetaOneBot$
rules become simpler.
Secondly, because we have decomposed $\tm{\cpCut{x}{P}{Q}}$ into an independent
name restriction and parallel composition, the $\kappa$-rules and the relevant
$\gamma$-rule all decompose as well.

@@

  Reductions are described by the smallest relation $\Longrightarrow$ on process
  terms closed under rules in \cref{fig:hcp-reduction}.

@@

\paragraph{Types}
We use the same definitions for types and environments for \hcp as we used for \cp.
However, we introduce a new layer on top of sequents: hypersequents.
As \cp is a one-sided logic, and it uses the left-hand side of the
turnstile to write the process, the traditional hypersequent notation can look
confusing: ``$\seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}$'' seems to claim
that $\tm{P}$ acts according to protocol $\ty{\Gamma_1}$.
What are all the other $\Gamma$s doing there? Are they typing empty processes?
Therefore, we opt to leave out the repeated turnstile, and instead work with the
notion of ``hyper-environments''. However, we will still refer to our system as a
hypersequent system.
Hyper-environments are either empty, or consist of a series of typing
environments, separated by vertical bars.
A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ types a series of
$n$ entangled, but independent processes.
\begin{definition}[Hyper-environment]\label{def:hcp-hyper-environment}
  \(
  \ty{\mathcal{G}}, \ty{\mathcal{H}} :=
  \ty{\emptyhypercontext}
  \; \mid \;
  \ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}
  \)
\end{definition}\noindent
Typing judgements in \hcp associate processes with hyper-environments.
\textsc{H-Mix} composes two processes in parallel, but remembers that they are
independent in the sequent.
\textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of
at least two independent processes, and connects them, eliminating the vertical
bar.
Note that each logical rule has the side condition that $\tm{x} \not\in
\ty{\mathcal{G}}$, which can be read as ``you cannot act on one end-point of $x$
if you are also holding its other end-point''.
This prevents self-locking processes, \eg
$\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes
  that the process $\tm{P}$ consists of $n$ independent, but potentially
  entangled processes, each of which communicates according to its own
  protocol $\Gamma_i$.
  Typing judgements can be constructed using the inference rules in \cref{fig:hcp}.
  Furthermore, each logical rule has the side condition that $\tm{x} \not\in
  \ty{\mathcal{G}}$.
\end{definition}\noindent
\begin{figure}[htb]
  \begin{center}\hccpInfAx \hccpInfCycle\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfMix \hccpInfHalt\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBoundTens \hccpInfParr \hccpInfOne\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBot \hccpInfPlus1 \hccpInfPlus2\end{center}
  \vspace*{0.5\baselineskip}
  \begin{center}\hccpInfWith \hccpInfNil \hccpInfTop\end{center}
  \caption{Hypersequent Classical Processes, typing.}
  \label{fig:hcp}
\end{figure}

@@

In this section, we discuss the relation between \cp and \hcp.
We will prove three important theorems:

@@

To facilitate this, we define a translation from terms in \cp to terms in \hcp
which breaks down the term constructs in \cp into their more atomic constructs
in \hcp.

@@

First, we prove that each \cp process can be translated by this trivial
translation to an \hcp process, and that this translation respects structural
congruence and reduction.
Reductions from \cp can be trivially translated to reductions in \hcp.
However, as we took apart each $\kappa$-rule into two separate rules, the
reduction relation of \hcp is a slight refinement over that of \cp.

@@

  By induction on the derivation of $\seq[P]{\Gamma}$.
  We show the interesting cases:

@@

  If $\reducesto{\mtf{P}}{R}$ in \hcp, then there is a $\tm{Q}$ such that
  $\reducesto{P}{Q}$ in \cp and $\tm{R}\Longrightarrow^\star\tm{\mtf{Q}}$ in \hcp.

@@

  By induction on the derivation of $\reducesto{\mtf{P}}{R}$.
  The cases for \hccpRedAxCut1 and the $\beta$-rules are trivial.
  For the $\kappa\nu$-rules, we rewrite by the appropriate $\kappa{\ppar}$ rule,
  and vice versa.

>>>>

\documentclass[draft,submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{example}{Example}
\input{preamble-typing}
\usepackage[sectionbib,square,numbers,sort&compress]{natbib}
\usepackage{more-space}
\usepackage{setspace}
\everypar{\looseness=-1}
\renewcommand{\baselinestretch}{0.975}

@@

This syntactic mismatch has an effect on the semantics as well. For instance, the $\beta$-reduction for output and input in \cp is \(\reducesto {\cpCut{x}{\cpSend{x}{y}{P}{Q}}{\cpRecv{x}{y}{R}}} {\cpCut{y}{P}{\cpCut{x}{Q}{R}}}\). Here, the parallel composition $\tm{(\piPar{P}{Q})}$ is of no relevance to this communication, yet the rule needs to inspect it to be able to nest the name restrictions appropriately in the resulting term.

@@

The term language of \cp is a variant of the \textpi-calculus. The variables $\tm{x}$, $\tm{y}$, and $\tm{z}$ range over channel names. The construct $\tm{\cpLink{x}{y}}$ links two channels~\cite{sangiorgi1996,boreale1998}, forwarding messages received on $\tm{x}$ to $\tm{y}$ and vice versa. The construct $\tm{\cpCut{x}{P}{Q}}$ creates a new channel $\tm{x}$, and composes two processes, $\tm{P}$ and $\tm{Q}$, which communicate on $\tm{x}$, in parallel. Therefore, in $\tm{\cpCut{x}{P}{Q}}$ the name $\tm{x}$ is bound in both $\tm{P}$ and $\tm{Q}$. In $\tm{\cpRecv{x}{y}{P}}$ and $\tm{\cpSend{x}{y}{P}{Q}}$, round brackets denote input, square brackets denote output. We use bound output~\cite{sangiorgi1996}, meaning that both input and output bind a new name. In $\tm{\cpRecv{x}{y}{P}}$ the new name $\tm{y}$ is bound in $\tm{P}$. In $\tm{\cpSend{x}{y}{P}{Q}}$, the new name $\tm{y}$ is only bound in $\tm{P}$, while $\tm{x}$ is only bound in $\tm{Q}$.

@@

Terms in \cp are identified up to structural congruence, which states that links are symmetric, and parallel compositions $\tm{\cpCut{x}{P}{Q}}$ are associative and commutative.

@@

Reductions relate processes with their reduced forms \eg a reduction $\reducesto{P}{Q}$ denotes that the process $\tm{P}$ can reduce to the process $\tm{Q}$ in a single step.

@@

  Reductions are described by the smallest relation $\Longrightarrow$ on process terms closed under rules in \cref{fig:cp-reduction}. The relation $\Longrightarrow^\star$ is the reflexive, transitive closure of 

@@

In this section, we introduce Hypersequent Classical Processes (\hcp), a variant of \cp which registers parallelism in the typing judgements using hypersequents, allowing us to take apart the monolithic term constructors of \cp (\eg $\tm{\cpSend{x}{y}{P}{Q}}$) into the corresponding \textpi-calculus term constructs.

@@

The term language of \hcp is a variant of \cp where the term constructs have been taken apart into \textpi-calculus primitives.

@@

One side-effect of the changes on the term level is that our structural congruence is much more standard. It has associativity, commutativity, and a unit for parallel composition, commutativity of name restrictions, and scope extrusion.

@@

  The structural congruence $\equiv$ is the congruence closure over terms which satisfies the following additional axioms:

@@

There are two changes to the reduction system. First, because $\tm{\piBoundSend{x}{y}{P}}$ and $\tm{\piBoundSend{x}{}{P}}$ are now terms in their own right, the $\cpRedBetaTensParr$ and $\cpRedBetaOneBot$ rules become simpler. Secondly, because we have decomposed $\tm{\cpCut{x}{P}{Q}}$ into an independent name restriction and parallel composition, the $\kappa$-rules and the relevant $\gamma$-rule all decompose as well.

@@

  Reductions are described by the smallest relation $\Longrightarrow$ on process terms closed under rules in \cref{fig:hcp-reduction}.

@@

\paragraph{Types}
We use the same definitions for types and environments for \hcp as we used for \cp. However, we introduce a new layer on top of sequents: hypersequents. As \cp is a one-sided logic, and it uses the left-hand side of the turnstile to write the process, the traditional hypersequent notation can look confusing: ``$\seq[P]{\Gamma_1}\hsep\dots\hsep\seq{\Gamma_n}$'' seems to claim that $\tm{P}$ acts according to protocol $\ty{\Gamma_1}$. What are all the other $\Gamma$s doing there? Are they typing empty processes? Therefore, we opt to leave out the repeated turnstile, and instead work with the notion of ``hyper-environments''. However, we will still refer to our system as a hypersequent system. Hyper-environments are either empty, or consist of a series of typing environments, separated by vertical bars. A hyper-environment $\ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ types a series of $n$ entangled, but independent processes.
\begin{definition}[Hyper-environment]\label{def:hcp-hyper-environment}
  $\ty{\mathcal{G}}, \ty{\mathcal{H}} := \ty{\emptyhypercontext} \; \mid \; \ty{\Gamma_1 \hsep \dots \hsep \Gamma_n}$
\end{definition}\noindent
Typing judgements in \hcp associate processes with hyper-environments. \textsc{H-Mix} composes two processes in parallel, but remembers that they are independent in the sequent. \textsc{H-Cut} and $(\tens)$ take as their premise a process which consists of at least two independent processes, and connects them, eliminating the vertical bar. Note that each logical rule has the side condition that $\tm{x} \not\in \ty{\mathcal{G}}$, which can be read as ``you cannot act on one end-point of $x$ if you are also holding its other end-point''. This prevents self-locking processes \eg $\tm{\piBoundSend{x}{}{\piRecv{x}{}{\piHalt}}}$.
\begin{definition}[Typing judgements]\label{def:hcp}
  A typing judgement $\seq[P]{\Gamma_1 \hsep \dots \hsep \Gamma_n}$ denotes that the process $\tm{P}$ consists of $n$ independent, but potentially entangled processes, each of which communicates according to its own protocol $\Gamma_i$. Typing judgements can be constructed using the inference rules in \cref{fig:hcp}. Furthermore, each logical rule has the side condition that $\tm{x} \not\in \ty{\mathcal{G}}$.
\end{definition}\noindent
\begin{figure}[htb]
  \begin{center}\hccpInfAx \hccpInfCycle\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfMix \hccpInfHalt\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBoundTens \hccpInfParr \hccpInfOne\end{center}
  \vspace*{-0.5\baselineskip}
  \begin{center}\hccpInfBot \hccpInfPlus1 \hccpInfPlus2\end{center}
  \vspace*{0.5\baselineskip}
  \begin{center}\hccpInfWith \hccpInfNil \hccpInfTop\end{center}
  \caption{Hypersequent Classical Processes, typing.}
  \label{fig:hcp}
\end{figure}

@@

In this section, we discuss the relation between \cp and \hcp. We will prove three important theorems:

@@

To facilitate this, we define a translation from terms in \cp to terms in \hcp which breaks down the term constructs in \cp into their more atomic constructs in \hcp.

@@

First, we prove that each \cp process can be translated by this trivial translation to an \hcp process, and that this translation respects structural congruence and reduction. Reductions from \cp can be trivially translated to reductions in \hcp. However, as we took apart each $\kappa$-rule into two separate rules, the reduction relation of \hcp is a slight refinement over that of \cp.

@@

  By induction on the derivation of $\seq[P]{\Gamma}$. We show the interesting cases:

@@

  If $\reducesto{\mtf{P}}{R}$ in \hcp, then there is a $\tm{Q}$ such that $\reducesto{P}{Q}$ in \cp and $\tm{R}\Longrightarrow^\star\tm{\mtf{Q}}$ in \hcp.

@@

  By induction on the derivation of $\reducesto{\mtf{P}}{R}$. The cases for \hccpRedAxCut1 and the $\beta$-rules are trivial. For the $\kappa\nu$-rules, we rewrite by the appropriate $\kappa{\ppar}$ rule, and vice versa.
